---
title: "Check lens shape data"
author: "Amartya Mitra, Molly Womack, Katie Thomas"
date: '2022-07-18'
output:
  html_document:
    code_fold: show
    theme: flatly
    toc: yes
    toc_float: yes
---

<style type="text/css">

body{ /* Normal  */
      font-size: 17px;
  }
  
</style>


```{r setup, include = FALSE}

# Markdown settings
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 6, fig.width = 8, attr.output='style="max-height: 400px;"') 

# Load packages
library(ape)
library(kableExtra)
library(cowplot)
library(tidyverse)
```


Here, we check various aspects of our lens shape data prior to testing our hypotheses. 

# Import data

```{r}
# Load tidy data
lens_data_tidy <- read.csv("../Data/Tidy/lens_data_tidy.csv", header = TRUE, na.strings=c("", "NA", " "))

#Load species means by life stage
species_stage_means <- read.csv("../Data/Tidy/species_stage_means.csv", header = TRUE, na.strings=c("", "NA", " "))
```


# Repeat measurements

The following dataframe subsets the specimens and lenses that were measured multiple times independently in order to determine the precision of measurement techniques. 

```{r subset to dataframe for repeat measures}

lens_repeat <- lens_data_tidy %>%
  filter(life_stage == "repeat" & specimen_id != "USNM299711")
  
```

# Coefficient of variation

The coefficient of variation (CV) is the ratio of the standard deviation to the mean. The higher the coefficient of variation, the greater the level of dispersion around the mean. Here, we compare the coefficients of variation for repeat measures within one specimen to the coefficients of variation across the full dataset. 

``` {r calculating cv}

lens_repeat %>%
  group_by(specimen_id) %>%
  summarise(coef.var.flat = (sd(flatness)/mean(flatness))*100,
            coef.var.aniso = (sd(anisotropy)/mean(anisotropy))*100,
            coef.var.spher = (sd(sphericity)/mean(sphericity))*100,
            coef.var.elong = (sd(elongation)/mean(elongation))*100)

# The coefficients of variation for all the measures are quite low < 10 %

# CVs calculated for the entire dataset show much higer variation so it is okay to use this method 
species_stage_means %>%
  group_by()%>%
  summarise(coef.var.flat = (sd(mean_flatness)/mean(mean_flatness))*100,
            coef.var.aniso = (sd(mean_anisotropy)/mean(mean_anisotropy))*100,
            coef.var.spher = (sd(mean_sphericity)/mean(mean_sphericity))*100,
            coef.var.elong = (sd(mean_elongation)/mean(mean_elongation))*100)

```

# Left lens vs. right lens

The following creates a dataframe with **one row per specimen**. Shape data for **each lens** are in **separate columns**. The dataframe is then split into two: one for left lenses and one for right lenses. These dataframes contain data for anurans only.

```{r left-right-wide}

# dataframe that averages by specimen
LR_lens_data <- lens_data_tidy %>%
  mutate(genus_species = as.factor(paste(genus, species, sep = "_"))) %>% # creates a genus_species column
  filter(life_stage != "repeat" & specimen_id != "BMNH1906.8.30.62" & life_stage != "artificial") %>%
  mutate_if(is.character, as.factor) %>% 
  group_by(specimen_id, genus_species, lens, life_stage, adult_ecology, activity_period) %>%
  summarise(mean_flatness = mean(flatness, na.rm = TRUE),
            mean_anisotropy = mean(anisotropy, na.rm = TRUE),
            mean_elongation = mean(elongation, na.rm = TRUE),
            mean_sphericity = mean(sphericity, na.rm = TRUE),
            mean_svl = mean(sv_length_mm, na.rm = TRUE),
            n = n()) %>%
  ungroup() 

#subset left and right data  then merge left and right by specimen
#arrange(LR_lens_data, specimen_id, genus_species, life_stage)

L_lens_data <- filter(LR_lens_data, lens == "left")
R_lens_data <- filter(LR_lens_data, lens == "right")

LR_lens_data <- full_join(L_lens_data, R_lens_data, 
                          by = c("specimen_id", "life_stage"),
                          suffix = c(".left", ".right"))

LR_lens_data<-as.data.frame(LR_lens_data)
row.names(LR_lens_data)<-paste(LR_lens_data$specimen_id,row.names(LR_lens_data),sep="_")

#remove any rows with missing left or right lens shape data
LR_lens_data <- filter(LR_lens_data, mean_flatness.left > 0 & mean_anisotropy.left > 0 & mean_elongation.left > 0 & mean_sphericity.left > 0)
LR_lens_data <- filter(LR_lens_data, mean_flatness.right > 0 & mean_anisotropy.right > 0 & mean_elongation.right > 0 & mean_sphericity.right > 0)

```

Next, we plot data for left lens against right lens for all 3D shape metrics

```{r left vs right lens shape}

# left v right for individual metrics

#create plots
flat_lr_plot <- ggplot(LR_lens_data, aes(y=mean_flatness.left, x=mean_flatness.right)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  ylab("Left lens flatness") +
  xlab("Right lens flatness") +
  geom_smooth(method = "lm")

anis_lr_plot <- ggplot(LR_lens_data, aes(y=mean_anisotropy.left, x=mean_anisotropy.right)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  ylab("Left lens anisotropy") +
  xlab("Right lens anisotropy") +
  geom_smooth(method = "lm")
  
elong_lr_plot <- ggplot(LR_lens_data, aes(y=mean_elongation.left, x=mean_elongation.right)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  ylab("Left lens elongation") +
  xlab("Right lens elongation") +
  geom_smooth(method = "lm")

spher_lr_plot <- ggplot(LR_lens_data, aes(y=mean_sphericity.left, x=mean_sphericity.right)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  ylab("Left lens sphericity") +
  xlab("Right lens sphericity") +
  geom_smooth(method = "lm")

#make figure
plot <- plot_grid(anis_lr_plot, flat_lr_plot, spher_lr_plot, elong_lr_plot,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A","B","C","D"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 2) #number of rows in grids

#print figure
plot
```

```{r, results="hide"}
#export figure
pdf(file = "../Outputs/Figures/Fig-supp-LR_lens_shape.pdf", width = 9,  height = 7)
plot
dev.off()
```


```{r}
# linear models of left vs right values for each lens metric
LR_flat<-lm(LR_lens_data$mean_flatness.left~LR_lens_data$mean_flatness.right)
LR_anis<-lm(LR_lens_data$mean_anisotropy.left~LR_lens_data$mean_anisotropy.right)
LR_elon<-lm(LR_lens_data$mean_elongation.left~LR_lens_data$mean_elongation.right)
LR_sphe<-lm(LR_lens_data$mean_sphericity.left~LR_lens_data$mean_sphericity.right)

#summary of each model
summary(LR_flat)
summary(LR_anis)
summary(LR_elon)
summary(LR_sphe)
```

We also examine model residuals from these regressions. 

```{r}

# naming the residuals of the models as specimen_id
names(LR_flat$residuals)<-LR_lens_data$specimen_id
names(LR_anis$residuals)<-LR_lens_data$specimen_id
names(LR_elon$residuals)<-LR_lens_data$specimen_id
names(LR_sphe$residuals)<-LR_lens_data$specimen_id

#plot ordered residuals

#pdf(file = "LeftVsRight_LensShape_OrderedLMResiduals_.pdf", width = 9,  height = 7)
par(mfrow=c(2, 2))

LR_flat$residuals <- sort(-(LR_flat$residuals))
plot(LR_flat$residuals)
abline(0,0, col="red")

LR_anis$residuals <- sort(-(LR_anis$residuals))
plot(LR_anis$residuals)
abline(0,0, col="red")

LR_elon$residuals <- sort(-(LR_elon$residuals))
plot(LR_elon$residuals)
abline(0,0, col="red")

LR_sphe$residuals <- sort(-(LR_sphe$residuals))
plot(LR_sphe$residuals)
abline(0,0, col="red")
#dev.off()

#plot ordered residuals with specimen labels
pdf(file = "../Outputs/LeftVsRight_LensShape_OrderedLMResiduals_Labeled_.pdf", width = 30,  height = 30)
par(mfrow=c(2, 2))
plot(LR_flat$residuals)
text(LR_flat$residuals, label=names(LR_anis$residuals))
plot(LR_anis$residuals)
text(LR_anis$residuals, label=names(LR_anis$residual))
plot(LR_elon$residuals)
text(LR_elon$residuals, label=names(LR_elon$residual))
plot(LR_sphe$residuals)
text(LR_sphe$residuals, label=names(LR_sphe$residual))
dev.off()

# Start writing to an output file
sink('../Outputs/LR_ShapeAssymetry_Outliers.txt')

#check for outliers
cat("\n\n LR_Flat_Outliers \n")
boxplot.stats(LR_flat$residuals)$out  
cat("\n\n LR_Anis_Outliers \n")
boxplot.stats(LR_anis$residuals)$out  
cat("\n\n LR_Elon_Outliers \n")
boxplot.stats(LR_elon$residuals)$out 
cat("\n\n LR_Sphe_Outliers \n")
boxplot.stats(LR_sphe$residuals)$out  

# Stop writing to the file
sink()

```

# Tests for correlation between 3D shape metrics

Correlation between all metrics. Flatness is excluded from further analyses to avoid multicollinearity because it is highly correlated with anisotropy 

```{r correlation in lens metrics}

#linear regressions and plots of pairs of shape metrics

#flatness vs. anisotropy
flat_anis <- ggplot(species_stage_means, aes(x = mean_anisotropy, y = mean_flatness, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Anisotropy", y="Flatness") +
  theme(legend.position = "none")

summary(lm(mean_flatness~mean_anisotropy, data = species_stage_means))


#flatness vs. elongation 
flat_elon <- ggplot(species_stage_means, aes(x = mean_elongation, y = mean_flatness, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Elongation", y="Flatness") +
  theme(legend.position = "none")

summary(lm(mean_flatness~mean_elongation, data = species_stage_means))


#flatness vs. sphericity
flat_sph <- ggplot(species_stage_means, aes(x = mean_sphericity, y = mean_flatness, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Sphericity", y="Flatness") +
  theme(legend.position = "none")

summary(lm(mean_flatness~mean_sphericity, data = species_stage_means))

#elongation vs. anisotropy
elong_anis <- ggplot(species_stage_means, aes(x = mean_anisotropy, y = mean_elongation, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Anisotropy", y="Elongation") +
  theme(legend.position = "none")

summary(lm(mean_elongation~mean_anisotropy, data = species_stage_means))


#elongation vs. sphericity
elong_sph <- ggplot(species_stage_means, aes(x = mean_sphericity, y = mean_elongation, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Sphericity", y="Elongation") +
  theme(legend.position = "none")

summary(lm(mean_elongation~mean_sphericity, data = species_stage_means))


#sphericity vs. anisotropy
sph_anis <- ggplot(species_stage_means, aes(x = mean_anisotropy, y = mean_sphericity, shape = life_stage)) +
  scale_shape_manual(values=c(19,1), name = "Life stage") +
  geom_point(alpha = 0.6) +
  theme_classic() +
  labs(x="Anisotropy", y="Sphericity") +
  theme(legend.position = "none")

summary(lm(mean_sphericity~mean_anisotropy, data = species_stage_means))

#construct supplemental figure
plots <- plot_grid(flat_anis, flat_elon, flat_sph, elong_anis, elong_sph, sph_anis,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A","B","C","D","E","F"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 3) #number of rows in grids

#print figure
plots

#export figure
pdf(file = "../Outputs/Figures/Fig-supp_shape_correlations.pdf", width = 10,  height = 12)
plots
dev.off()
```

# Tests for correlation between lens shape and body size

We checked that among adults that we had body size (snout-vent length) data for, there was no correlation between body size and any of the 3D shape metrics, as this could influence our models and interpretations. 


```{r snout-vent length and shape}

#regressions of lens shape vs. SVL 
summary(lm(mean_sphericity~mean_svl, data = species_stage_means))
summary(lm(mean_anisotropy~mean_svl, data = species_stage_means))
summary(lm(mean_flatness~mean_svl, data = species_stage_means))
summary(lm(mean_elongation~mean_svl, data = species_stage_means))

#plots of lens shape by body size
par(mfrow=c(2, 2))
plot(mean_sphericity~mean_svl, data = species_stage_means)
plot(mean_anisotropy~mean_svl, data = species_stage_means)
plot(mean_flatness~mean_svl, data = species_stage_means)
plot(mean_elongation~mean_svl, data = species_stage_means)

#plot of mean body size across adult habitats
par(mfrow=c(1, 1))
plot(mean_svl~as.factor(adult_ecology), data = species_stage_means)

```
