---
title: "Lens shape in tadpoles vs. adults"
author: "Amartya Mitra, Molly Womack, Katie Thomas"
date: '2022-07-21'
output:
  html_document:
    code_fold: show
    theme: flatly
    toc: yes
    toc_float: yes
---

<style type="text/css">

body{ /* Normal  */
      font-size: 17px;
  }
  
</style>


```{r setup, include = FALSE}

# Markdown settings
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 6, fig.width = 8, attr.output='style="max-height: 400px;"') 

# Load packages
library(ape)
library(kableExtra)
library(gridExtra)
library(cowplot)
library(tidyverse)
library(geomorph)
library(caper)
library(geiger)
```


Here, we test whether tadpole lens shape differs from adult lens shape

# Import data

```{r}
# Load tidy data
lens_data_tidy <- read.csv("../Data/Tidy/lens_data_tidy.csv", header = TRUE, na.strings=c("", "NA", " "))

#Load species means by life stage
species_stage_means <- read.csv("../Data/Tidy/species_stage_means.csv", header = TRUE, na.strings=c("", "NA", " "))

#Load data for specimen means
specimen_means <- read.csv("../Data/Tidy/specimen_means.csv", header = TRUE, na.strings=c("", "NA", " "))


#Import phylogeny from Jetz and Pyron 2019
tree.full <- read.nexus(file = "../Data/Tidy/lens_shape_tree.nex") 
```

## Q1 Lens shape of tadpoles vs. adults

Here, we examine whether tadpole lens shape differs from adult lens shape across all species sampled (118 species with adult data, 50 species with tadpole data). These models do not include phylogenetic corrections. 

### Principle component analysis

Here, we run a PCA on species means by life stage for all tadpoles and adults sampled. Shape data included are anisotropy, elongation, and sphericity. 

```{r PCA}

#dimension conversion
row.names(species_stage_means) <- paste(species_stage_means$life_stage, row.names(species_stage_means), sep="_") 
species_stage_means$life_Stage <- NULL

# running PCA on tadpoles and adults with anisotropy, elongation and sphericity values
lens.pca <- prcomp(species_stage_means[,c("mean_anisotropy","mean_elongation","mean_sphericity")], center = TRUE, scale. = TRUE)

# summary stats
summary(lens.pca)

str(lens.pca)

# eigenvectors
lens.pca$rotation

# eigenvalues 
(lens.pca$sdev)^2

# plot directly
plot(lens.pca$x[,1], lens.pca$x[,2])
```


### t-tests 

Here, we use Welch's two-sample t tests to test whether tadpoles and adults exhibit different lens shapes for each of the three shape metrics examined (anisotropy, sphericity, and elongation).

```{r independent t-test all tads and adults}

# unequal variances therefore welch two sample t-test 

#anisotropy
t.test(mean_anisotropy~life_stage, data = species_stage_means)

#sphericity
t.test(mean_sphericity~life_stage, data = species_stage_means)

#elongation
t.test(mean_elongation~life_stage, data = species_stage_means)
```

### Procrustes MANOVA for shape variables

Here, we use a multivariate approach to test whether the three shape metrics together (sphericity, anisotropy, elongation) are associated with life stage (tadpole vs. adult)

```{r procD all metrics, eval = TRUE}

fit <- procD.lm(species_stage_means[,c("mean_anisotropy","mean_elongation","mean_sphericity")]~life_stage,iter = 999,seed = NULL,RRPP = TRUE, SS.type = NULL, int.first = FALSE, Cov = NULL, data = species_stage_means, print.progress = TRUE)

summary(fit)
```

## Q2 Differences in lens shape through ontogeny within species

Here, we examine a subset of species (n = 45) that have matched tadpole and adult lens shape data to see how lens shape changes through ontogeny and whether this is driven by differences in adult habitat. 

### Phylogenetic signal

Here, we estimated phylogenetic signal in tadpole lens shape, in adult lens shape, and in the change in lens shape through ontogeny (difference between adults and tadpoles within species). For these analyses, we used the physignal function in the geomorph package, which estimates the multivariate version of the K-statistic (Kmult; Adams 2014) and allowed us to incorporate all three shape metrics (anisotropy, sphericity, elongation) simultaneously. 

```{r phylogenetic signal, fig.height = 12, fig.width = 5}

#Prepare data -----

# PCA with all data tadpole Xs and adults Os
pca_data <- cbind(as.data.frame(species_stage_means), as.data.frame(lens.pca$x))

pca_means <- pca_data %>%
  ungroup() 

#make separate dataframes for adults and tadpoles
phylo_tads <- filter(pca_means, life_stage == "tadpole")
phylo_adults <- filter(pca_means, life_stage == "adult")

#Merge dataset by genus species
phylo_pca_means <- full_join(phylo_tads, phylo_adults, 
                             by = "genus_species", 
                             suffix = c(".tads", ".adults")) %>%
  rename(adult_ecology = adult_ecology.adults) %>%
  rename(activity_period = activity_period.adults) %>%
  mutate(family = ifelse(is.na(family.adults), as.character(family.tads), as.character(family.adults))) %>%
  dplyr::select(genus_species, family, mean_flatness.tads, mean_anisotropy.tads, mean_elongation.tads, mean_sphericity.tads, mean_svl.tads, PC1.tads, PC2.tads, PC3.tads, n_specimens.tads, mean_flatness.adults, mean_anisotropy.adults, mean_elongation.adults, mean_sphericity.adults, mean_svl.adults, PC1.adults, PC2.adults, PC3.adults, n_specimens.adults, adult_ecology, activity_period)

#export data subset
write.csv(phylo_pca_means, file = "../Data/Tidy/phylo_pca_means.csv", row.names = FALSE)

#Build tadpole, adult, and matched species datasets-------

#tadpole data
tad_phylo_means <- filter(phylo_pca_means, mean_flatness.tads > 0)

#adult data
adult_phylo_means <- filter(phylo_pca_means, mean_flatness.adults > 0)

#matched data (both)
both_phylo_means <- filter(tad_phylo_means, mean_flatness.adults > 0)
both_phylo_means$mean_flatness.diff<-both_phylo_means$mean_flatness.tads-both_phylo_means$mean_flatness.adults
both_phylo_means$mean_anisotropy.diff<-both_phylo_means$mean_anisotropy.tads-both_phylo_means$mean_anisotropy.adults
both_phylo_means$mean_elongation.diff<-both_phylo_means$mean_elongation.tads-both_phylo_means$mean_elongation.adults
both_phylo_means$mean_sphericity.diff<-both_phylo_means$mean_sphericity.tads-both_phylo_means$mean_sphericity.adults

# Build tadpole, adult, and matched data trees -----

#check tree for all species
rownames(phylo_pca_means) <- phylo_pca_means$genus_species;
name.check(tree.full, phylo_pca_means)->overlap
overlap
#plot(tree.full)

#trim tree to tad data
rownames(tad_phylo_means) <- tad_phylo_means$genus_species
name.check(tree.full, tad_phylo_means) -> overlap
#overlap
tree.tad <- drop.tip(tree.full, overlap$tree_not_data)
#plot(tree.tad)

#export tadpole species tree
write.nexus(tree.tad, file = "../Data/Tidy/tree_tadpole.nex")

#sort tad data to match tad tree
rownames(tad_phylo_means) <- tad_phylo_means$genus_species
tad_phylo_means <- tad_phylo_means[tree.tad$tip.label, ]

#trim tree to adult data
rownames(adult_phylo_means) <- adult_phylo_means$genus_species
name.check(tree.full,adult_phylo_means)->overlap
#overlap
tree.adult <- drop.tip(tree.full,overlap$tree_not_data)
#plot(tree.adult)

#export adult species tree
write.nexus(tree.adult, file = "../Data/Tidy/tree_adult.nex")

#sort adult data to match adult tree
rownames(adult_phylo_means) <- adult_phylo_means$genus_species
adult_phylo_means <- adult_phylo_means[tree.adult$tip.label, ]

#trim tree to species with matched tad and adult (both) data
rownames(both_phylo_means) <- both_phylo_means$genus_species;
name.check(tree.full, both_phylo_means) -> overlap
#overlap
drop.tip(tree.full, overlap$tree_not_data) -> tree.both;
#plot(tree.both)

#reorder matched data to follow matched tree
rownames(both_phylo_means) <- both_phylo_means$genus_species;
both_phylo_means <-  both_phylo_means[tree.both$tip.label, ]

# Tests for phylogenetic signal ------

#phylogenetic signal of tadpole lens shape (all species of tads, n=48)
physignal(as.matrix(tad_phylo_means[,c("mean_anisotropy.tads","mean_elongation.tads","mean_sphericity.tads")]), tree.tad,iter = 999, seed = NULL)

#phylogenetic signal of adult lens shape (all species of adults, n=118)
physignal(as.matrix(adult_phylo_means[,c("mean_anisotropy.adults","mean_elongation.adults","mean_sphericity.adults")]), tree.adult, iter = 999, seed = NULL)

#phylogenetic signal of the difference between tad and adult lens shape (n=45 species)
physignal(as.matrix(both_phylo_means[,c("mean_anisotropy.diff","mean_elongation.diff","mean_sphericity.diff")]), tree.both, iter = 999, seed = NULL)

#check phy signal for tadpoles in matched species only (n=45 species)
physignal(as.matrix(both_phylo_means[,c("mean_anisotropy.tads","mean_elongation.tads","mean_sphericity.tads")]), tree.both, iter = 999, seed = NULL)

#check phy signal for adults in matched species only (n=45 species)
physignal(as.matrix(both_phylo_means[,c("mean_anisotropy.adults","mean_elongation.adults","mean_sphericity.adults")]), tree.both, iter = 999, seed = NULL)

```

### Difference between tadpole and adult lens shape vs ecology

Here, we examine whether the change in lens shape through ontogeny (difference between tadpole and adult lens shape within species) differs across species that remain aquatic vs. transition to terrestrial environments. We do this first by examining each 3D lens metric individually, and then with a multivariate approach.

#### PGLS 

```{r}

#pgls regressions testing for effect of adult habitat on each 3D shape metric (separately)

#prep data/tree ----

#add aquatic vs. terrestrial category
both_phylo_means$env <- ifelse(both_phylo_means$adult_ecology=="aquatic", "aquatic", "terrestrial")

#align data and phylogeny labels
both_phylo_means <- both_phylo_means[tree.both$tip.label, ] 

#label tree nodes
geocapertree <- makeNodeLabel(tree.both, method = "number", prefix = "Node")

#make comparative object
geocaperdata <- comparative.data(phy=geocapertree, data=both_phylo_means, names.col=genus_species, vcv=TRUE, na.omit=FALSE, warn.dropped=TRUE)

#pgls for anisotropy----

#fit model
aniso.pgls <- pgls(mean_anisotropy.diff~env, data=geocaperdata, lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(aniso.pgls)
par(mfrow = c(1, 1))

#view coefficients
summary(aniso.pgls)


#pgls for sphericity----

#fit model
sphere.pgls <- pgls(mean_sphericity.diff~env, data=geocaperdata, lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(sphere.pgls)
par(mfrow = c(1, 1))

#view coefficients
summary(sphere.pgls)


#pgls for elongation----

#fit model
elong.pgls<-pgls(mean_elongation.diff~env,data=geocaperdata, lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(elong.pgls)
par(mfrow = c(1, 1))

#view coefficients
summary(elong.pgls)
```

#### Procrustes MANOVA

```{r, eval = TRUE}

#check adequate levels for testing
table(both_phylo_means$env)

#procD.pgls of lens shape differences in all 3 measures 
procD <- geomorph::procD.pgls(both_phylo_means[,c("mean_anisotropy.diff","mean_elongation.diff","mean_sphericity.diff")] ~ env, tree.both, Cov = NULL, iter = 999,seed = NULL, int.first = FALSE, SS.type = NULL, effect.type = NULL, data = both_phylo_means, print.progress = TRUE)

#model summary
summary(procD)
```

# Figures

```{r}
#define a colorblind-friendly vector of colors for adult habitat
col_hab <- c("aquatic" = "#0072B2",
              "fossorial" = "#D55E00",
              "ground-dwelling" = "#E69F00",
              "scansorial" = "#009E73",
              "semi-aquatic" = "#56B4E9",
              "subfossorial" = "#CC79A7")

#create vector of colors for activity period
col_act <- c("both" = "deeppink3",
             "diurnal" = "darkgoldenrod1", 
             "nocturnal" = "blueviolet")

#create dataframe of artifical lens shapes
lens_sh <- specimen_means %>%
  filter(life_stage=="artificial") %>%
  add_row(specimen_id = "Perfect sphere", mean_anisotropy = 0, mean_flatness = 1, mean_sphericity = 1, mean_elongation = 1) %>%
  filter(specimen_id != "Sphere")
```

## Figure 2 : Comparison of tadpole and adult lens shapes

Panel A - PCA of tadpoles vs adults across all specimens, Panel B - Violin plots of species means for 3 shape metrics by tadpole and adult life stages

```{r Figure 2, fig.width=6, fig.height=8}

#### Panel A: tadpole and adult lens shape PCA ####

#put PCA results into dataframe format
df_pca <- as.data.frame(lens.pca$x) 

#add a group column by breaking apart the rownames from the original dataframe
df_pca$group <- sapply(strsplit(as.character(row.names(species_stage_means)), "_"),  "[", 1)

#alternate attempt to get life stage to merge with df_pca
## is pca dataframe in same order as species_stage_mean dataframe?? if not, this is not correct.
df_pca$group <- species_stage_means$life_stage

#extract loadings of the variables
pca_loadings <- data.frame(Variables = rownames(lens.pca$rotation), lens.pca$rotation)

#make ggplot of results  
plot_pca <- ggplot(df_pca, 
                   aes(x = PC1, y = PC2)) +
  geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = (PC1*4), # adds arrows - eigenvectors
                                        yend = (PC2*4)), arrow = arrow(length = unit(1/2, "picas")),
               color = "black") +
  #annotate("text", x = (pca_loadings$PC1*4), y = (pca_loadings$PC2*4), label = pca_loadings$Variables) + #adds eigenvector labels. Not using in this case because too hard to position
  stat_ellipse(data = filter(df_pca, group == "tadpole"), aes(x=PC1, y=PC2), type = "norm", col="gray") + # adds elliptical cluster for tadpole group
  stat_ellipse(data = filter(df_pca, group == "adult"), aes(x=PC1, y=PC2), type = "norm") + # adds elliptical cluster for adult group
  geom_point(aes(fill=group), pch = 21, alpha = 0.8) + 
  scale_fill_manual(values=c("black","white"), name = "Life stage") +
  xlab("PC 1 (53%)") + 
  ylab("PC 2 (33%)") +
  theme_classic()+
  theme(panel.grid = element_blank(), 
        panel.border = element_blank(),
        legend.position = "top",
        plot.margin = margin(l = 15, unit = "pt"))

plot_pca

#note for the arrows, the one pointing left is for anisotropy, up is for flatness, to the right is for sphericity. 

#### Panel B: violin of 3 shape metrics by tadpole and adult ####

#make 3-panel plot for tad/adult shapes

#put specimen data into long format for plotting
value <- species_stage_means$mean_anisotropy
stage <- species_stage_means$life_stage
metric <- "anisotropy"
shape1 <- data.frame(metric, value, stage)

value <- species_stage_means$mean_sphericity
stage <- species_stage_means$life_stage
metric <- "sphericity"
shape3 <- data.frame(metric, value, stage)

value <- species_stage_means$mean_elongation
stage <- species_stage_means$life_stage
metric <- "elongation"
shape4 <- data.frame(metric, value, stage)

shape.final <- rbind(shape1, shape3, shape4)

#anisotropy plot
anis.plot <- shape.final %>%
  filter(metric == "anisotropy") %>%
  ggplot(aes(y = value, x = stage)) +
  scale_fill_manual(values=c("black", "white")) +
  geom_point(aes(fill = stage), pch = 21, alpha = 0.6, size = .7,  position=position_jitterdodge(dodge.width=1, jitter.width = 1)) +
  geom_violin(alpha=0.1, trim = T, width = 1, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), alpha=0.1, outlier.shape = NA) + #controls boxes
  theme_classic()+
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = 'none') + #controls background
  xlab("") +
  ylab("Anisotropy") +
  ylim(c(0,0.5)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_anisotropy, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")


#sphericity plot
spher.plot <- shape.final %>%
  filter(metric == "sphericity") %>%
 ggplot(aes(y = value, x = stage)) +
  scale_fill_manual(values=c("black", "white"), guide = "none") +
  geom_point(aes(fill = stage), pch = 21, alpha = 0.6, size = .7,  position=position_jitterdodge(dodge.width=1, jitter.width = 1)) +
  geom_violin(alpha=0.1, trim = T, width = 1, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), alpha=0.1) + 
  theme_classic()+
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = 'none') + 
  xlab("") +
  ylab("Sphericity") +
  ylim(c(0.75,1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_sphericity, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
 
#elongation value
elong.plot <- shape.final %>%
  filter(metric == "elongation") %>%
  ggplot(aes(y = value, x = stage)) +
  scale_fill_manual(values=c("black", "white")) +
  geom_point(aes(fill = stage), pch = 21, alpha = 0.6, size = .7,  position=position_jitterdodge(dodge.width=1, jitter.width = 1)) +
  geom_violin(alpha=0.1, trim = T, width = 1, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), alpha=0.1) + #controls boxes
  theme_classic()+
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = 'none') + #controls background
  xlab("") +
  ylab("Elongation") +
  ylim(c(0.75,1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_elongation, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dotdash","dotdash"),#added dotdash because elongation for both shapes identical, can't see that they are both plotted so adding a dotdash that combines them
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes") 

# Construct figure -------

#label pca panel
plot_top <- plot_grid(plot_pca,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 1) #number of rows in grids


# arrange 3 shape panels together
plots_bottom <- plot_grid(anis.plot, spher.plot, elong.plot,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("B", "C", "D"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 1) #number of rows in grids

#make legend
leg <- get_legend(spher.plot + theme(legend.position="bottom"))

#combine plots and legend
plots <- plot_grid(plot_top, plots_bottom, leg,
                   align = 'vh',
                   axis = 'lb',
                   nrow = 3,
                   rel_heights = c(1, 0.9, 0.1))

#show figure 
plots

#export figure
pdf(file = "../Outputs/Figures/Fig2.pdf", width = 6,  height = 8)
plots
dev.off()
```


## Figure 3A : Phylogeny with PC scores. 

Species names are coloured by microhabitat. Circles and dotted lines added in illustrator

```{r Figure 3, fig.height = 12, fig.width = 6}

# Extracting PC scores
PC1_tad<-phylo_pca_means$PC1.tads
names(PC1_tad)<-phylo_pca_means$genus_species
PC1_tad<-PC1_tad[tree.full$tip.label]

PC1_adult<-phylo_pca_means$PC1.adults
names(PC1_adult)<-phylo_pca_means$genus_species
PC1_adult<-PC1_adult[tree.full$tip.label]


# Plot tree
plot(tree.full, show.tip.label= T, font=3, cex=0.5, label.offset = 1.5, edge.width=3)
segments(340 + (PC1_tad*40), 1:nrow(phylo_pca_means), 340 + (PC1_adult*10), 1:nrow(phylo_pca_means), col="grey90", lwd=5)
segments(340 + (PC1_tad*40), 1:nrow(phylo_pca_means), 340 + (PC1_tad*40), 1:nrow(phylo_pca_means), col="grey", lwd=5)
segments(340 + (PC1_adult*10), 1:nrow(phylo_pca_means), 340 + (PC1_adult*10), 1:nrow(phylo_pca_means), col="black", lwd=5)

lastPP<-get("last_plot.phylo",env=.PlotPhyloEnv)

SpEcology<-as.factor(phylo_pca_means$adult_ecology)
SemiEcoColr<-c("#0072B2","#D55E00","#E69F00","#009E73","#56B4E9","#CC79A7","black")
names(SemiEcoColr) <- levels(SpEcology)
SemiEcoColr<- SemiEcoColr[match(SpEcology, names(SemiEcoColr))] # col.gp must NOT be a factor
names(SemiEcoColr)<-phylo_pca_means$genus_species
SemiEcoColr<-SemiEcoColr[tree.full$tip.label]


#Print graph
plot(tree.full, show.tip.label= F, font=3, cex=0.5, label.offset = 1.5, edge.width=3, x.lim=c(0,700))
par(fg="black")
tt<-gsub("_"," ",tree.full$tip.label)
#plot segments for connecting tip labels to tad lens
segments(300, 1:nrow(phylo_pca_means), 550 + (pmax(PC1_adult, PC1_tad, na.rm=TRUE)*50), 1:nrow(phylo_pca_means), col="black", lwd=1, lty = "dotted")
#plot tip labels
text(lastPP$xx[1:length(tt)],lastPP$yy[1:length(tt)], tt, cex=0.5,pos=4, offset=0.1, font=3, col = "black")
#plot segments for change in lens
segments(550 + (PC1_tad*50), 1:nrow(phylo_pca_means), 550 + (PC1_adult*50), 1:nrow(phylo_pca_means), col="grey90", lwd=6)
#plot tadpole lens
points(550 + (PC1_tad*50), 1:nrow(phylo_pca_means), col="black", bg = "white", pch = 21, cex=1)
#plot adult lens
points(550 + (PC1_adult*50), 1:nrow(phylo_pca_means), col=SemiEcoColr, bg=SemiEcoColr, pch = 21, cex=1)


#Export figure
pdf(file = "../Outputs/Figures/Fig3a.pdf", width = 6,  height = 12)
plot(tree.full, show.tip.label= F, font=3, cex=0.5, label.offset = 1.5, edge.width=1.5, x.lim=c(0,700))
par(fg="black")
tt<-gsub("_"," ",tree.full$tip.label)
#plot segments for connecting tip labels to tad lens
segments(410, 1:nrow(phylo_pca_means), 575 + (pmax(PC1_adult, PC1_tad, na.rm=TRUE)*50), 1:nrow(phylo_pca_means), col="black", lwd=1, lty = "dotted")
#plot tip labels
text(lastPP$xx[1:length(tt)],lastPP$yy[1:length(tt)], tt, cex=0.5,pos=4, offset=0.1, font=3, col = "black")
#plot segments for change in lens
segments(575 + (PC1_tad*50), 1:nrow(phylo_pca_means), 575 + (PC1_adult*50), 1:nrow(phylo_pca_means), col="grey90", lwd=6)
#plot tadpole lens
points(575 + (PC1_tad*50), 1:nrow(phylo_pca_means), col="black", bg = "white", pch = 21, cex=1)
#plot adult lens
points(575 + (PC1_adult*50), 1:nrow(phylo_pca_means), col=SemiEcoColr, bg=SemiEcoColr, pch = 21, cex=1)
dev.off()
```

## Figure 3BC & S4: Species with matched tadpole and adult lens shape data across adult habitats. 

```{r Figure 4}

# Prep data for tadpole and adult lens shape within species --- 

#put data in long format

#tadpole data
tadData <- data.frame(species = both_phylo_means$genus_species,
                      anisotropy = both_phylo_means$mean_anisotropy.tads,
                      sphericity = both_phylo_means$mean_sphericity.tads,
                      elongation = both_phylo_means$mean_elongation.tads,
                      eco = both_phylo_means$adult_ecology,
                      stage = "tadpole")

#adult data
adultData <- data.frame(species = both_phylo_means$genus_species,
                        anisotropy = both_phylo_means$mean_anisotropy.adults,
                        sphericity = both_phylo_means$mean_sphericity.adults,
                        elongation = both_phylo_means$mean_elongation.adults,
                        eco = both_phylo_means$adult_ecology,
                        stage = "adult")

#combine datasets
matchData <- rbind(tadData, adultData)
matchData$stage <- factor(matchData$stage, levels = c("tadpole", "adult"))
matchData$eco <- factor(matchData$eco, levels = c("aquatic","fossorial","subfossorial","semi-aquatic","ground-dwelling","scansorial"))

#add column binning adults into "aquatic" or "terrestrial"
matchData$hab <- ifelse(matchData$eco=="aquatic", "aquatic", "terrestrial")

#make column that codes for both tadpole and adult habitat (for coloring)
matchData$hab2 <- ifelse(matchData$stage=="tadpole", "aquatic", matchData$hab)

#make column for faceting/labels based on adult habitat
matchData$lab <- ifelse(matchData$eco=="aquatic", "remains aquatic", "becomes terrestrial")
matchData$lab <- factor(matchData$lab, levels = c("remains aquatic", "becomes terrestrial"))

# Make Figure 4 ------

#define habitat colors
col_hab_binary <- c("aquatic" = "#0072B2",
                    "terrestrial" = "#4d3208")

#define life stage points
shape_stage <- c("tadpole" = 1,
                 "adult" = 19)

#plot matched data separated by adult habitats

#anisotropy
an.paired.plot2 <- ggplot(matchData, aes(x = stage, y = anisotropy)) +
  scale_fill_manual(values = col_hab_binary, name = "Habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab_binary, name = "Habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.6, aes(shape = stage)) + 
  geom_line(aes(group = species, col = hab), alpha = 0.4) +
  labs(x = "Life stage", y = "Anisotropy") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~lab, nrow = 1) +
  theme(strip.background = element_blank(), strip.text.x = element_text(size=9)) +
  geom_segment(aes(x=1, y = 0.4, xend = 2, yend = 0.4, col=hab))

#sphericity
sp.paired.plot2 <- ggplot(matchData, aes(x = stage, y = sphericity)) +
  scale_fill_manual(values = col_hab_binary, name = "Habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab_binary, name = "Habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.6, aes(shape = stage)) + 
  geom_line(aes(group = species, col = hab), alpha = 0.4) +
  labs(x = "Life stage", y = "Sphericity") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~lab, nrow = 1) +
  theme(strip.background = element_blank(), strip.text.x = element_text(size=10)) #+
  #geom_segment(aes(x=1, y = 1.01, xend = 2, yend = 1.01, col=hab))

#elongation
el.paired.plot2 <- ggplot(matchData, aes(x = stage, y = elongation)) +
  scale_fill_manual(values = col_hab_binary, name = "Habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab_binary, name = "Habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.6, aes(shape = stage)) +
  geom_line(aes(group = species, col = hab), alpha = 0.4) +
  labs(x = "Life stage", y = "Elongation") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~lab, nrow = 1) +
  theme(strip.background = element_blank(), strip.text.x = element_text(size=10)) +
  geom_segment(aes(x=1, y = 1.01, xend = 2, yend = 1.01, col=hab))

# Construct Figure 3BC -------

fig.b <- an.paired.plot2 +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) +
  geom_hline(data = lens_sh, aes(yintercept = mean_anisotropy, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")

fig.c <- sp.paired.plot2 +
  theme(strip.background = element_blank(), strip.text.x = element_blank()) +
  geom_hline(data = lens_sh, aes(yintercept = mean_sphericity, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")

plots <- plot_grid(fig.b, fig.c,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("B", "C"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 2) #number of rows in grids

#print figure
plots

#export figure
pdf(file = "../Outputs/Figures/Fig3bc.pdf", width = 3,  height = 8)
plots
dev.off()


# Make supplemental figure -----

#define a colorblind-friendly vector of colors for adult habitat
col_hab <- c("aquatic" = "#0072B2",
              "fossorial" = "#D55E00",
              "ground-dwelling" = "#E69F00",
              "scansorial" = "#009E73",
              "semi-aquatic" = "#56B4E9",
              "subfossorial" = "#CC79A7")

shape_stage <- c("tadpole" = 1,
                 "adult" = 19)

#plot matched data separated by adult habitats

#anisotropy
an.paired.plot <- ggplot(matchData, aes(x = stage, y = anisotropy, fill = eco)) +
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab, name = "Adult habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.9, aes(shape = stage)) + 
  geom_line(aes(group = species, col = eco), alpha = 0.5) +
  labs(x = "Life stage", y = "Anisotropy") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~eco, nrow = 1) +
  theme(strip.background = element_blank())
  
#sphericity
sp.paired.plot <- ggplot(matchData, aes(x = stage, y = sphericity, fill = eco)) +
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab, name = "Adult habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.9, aes(shape = stage)) + 
  geom_line(aes(group = species, col = eco), alpha = .5) +
  labs(x = "Life stage", y = "Sphericity") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~eco, nrow = 1)+
  theme(strip.background = element_blank())

#elongation
el.paired.plot <- ggplot(matchData, aes(x = stage, y = elongation, fill = eco)) +
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  scale_shape_manual(values = shape_stage, name = "Life stage") +
  scale_color_manual(values = col_hab, name = "Adult habitat") +
  geom_violin(alpha=0.5, trim = T, width = .8, position = position_dodge(1)) + 
  geom_boxplot(width = 0.2, position = position_dodge(1), fill="white", outlier.alpha = 0) +
  geom_point(size = 1, alpha = 0.9, aes(shape = stage)) + 
  geom_line(aes(group = species, col = eco), alpha = .5) +
  labs(x = "Life stage", y = "Elongation") + 
  theme(text = element_text(size=12), panel.background = element_blank(), axis.line = element_line(colour = "black"), legend.position = "none") +
  facet_wrap(~eco, nrow = 1)+
  theme(strip.background = element_blank())


# Construct supplemental figure -------

fig.a <- an.paired.plot +
  theme(axis.title.x=element_blank(), axis.text.x=element_blank()) 

fig.b <- sp.paired.plot +
  theme(strip.background = element_blank(), strip.text.x = element_blank(), axis.title.x=element_blank(), axis.text.x=element_blank()) 

fig.c <- el.paired.plot +
  theme(strip.background = element_blank(), strip.text.x = element_blank()) 


plots <- plot_grid(fig.a, fig.b, fig.c,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A", "B", "C"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 3) #number of rows in grids

#print figure
plots

#export figure
pdf(file = "../Outputs/Figures/FigS4.pdf", width = 8,  height = 11)
plots
dev.off()
```


