---
title: "Lens shape and relative size by adult ecology"
author: "Katie Thomas"
date: '2022-03-15'
output:
  html_document:
    code_fold: show
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

<style type="text/css">

body{ /* Normal  */
      font-size: 17px;
  }
  
</style>


```{r setup, include = FALSE}

# Markdown settings
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 6, fig.width = 8, attr.output='style="max-height: 400px;"') 

library(cowplot)
library(tcltk)
library(ape)
library(phytools)
library(geiger)
library(stringr)
library(kableExtra)
library(plotly)
library(RRPP)
library(viridis)
library(pca3d)
library(gridExtra)
library(geomorph)
library(caper)
library(tidyverse)
```

Here, we examine whether adult lens size (relative to eye size) or shape are correlated with adult ecology (habitat or activity period). We do this across all adults sampled with available data (n = 181 species, with n = 80 for eye size data)


# Import data

```{r}
# Load tidy data
phylo_pca_means <- read.csv("../Data/Tidy/phylo_pca_means.csv", header = TRUE, na.strings=c("", "NA", " "))

# Load eye data subset
adult_eye <- read.csv("../Data/Tidy/adult_eye.csv", header = TRUE, na.strings=c("", "NA", " "))

#Load species means by life state
species_stage_means <- read.csv("../Data/Tidy/species_stage_means.csv", header = TRUE, na.strings=c("", "NA", " "))

#Load specimen means
specimen_means <- read.csv("../Data/Tidy/specimen_means.csv", header = TRUE, na.strings=c("", "NA", " "))

#Import phylogeny from Jetz and Pyron 2019
tree.full <- read.nexus(file = "../Data/Tidy/lens_shape_tree.nex") 
```

Subset data by life stage

```{r}
#tadpole data
tad_phylo_means <- filter(phylo_pca_means, mean_flatness.tads > 0)

#adult data
adult_phylo_means <- filter(phylo_pca_means, mean_flatness.adults > 0)
```

# Adult lens shape vs. adult ecology 

## PGLS: Adult lens shape vs. ecology for each 3D shape metric

Here, we run a series of PGLS regressions to test for associations between each 3D shape metric (anisotropy, elongation, and sphericity) and adult ecology (habitat and activity period). 

### Habitat

```{r}
#pgls regressions testing for effect of adult habitat on each 3D shape metric (separately)

#prep data and tree----

#subset adults with habitat data
adult_phylo_eco<-filter(adult_phylo_means, !is.na(adult_ecology))
table(adult_phylo_eco$adult_ecology)

#trim tree to adults with habitat data
rownames(adult_phylo_eco)<-adult_phylo_eco$genus_species;
name.check(tree.full,adult_phylo_eco)->overlap
#overlap
drop.tip(tree.full, overlap$tree_not_data) -> tree.adult.eco;
#plot(tree.adult.eco)
#check tree and data match
name.check(tree.adult.eco, adult_phylo_eco)

#Remove excess data from species not found in tree
adult_phylo_eco<-adult_phylo_eco[ ! adult_phylo_eco$genus_species %in% overlap$data_not_tree, ]
adult_phylo_eco[] <- lapply(adult_phylo_eco, function(x) if(is.factor(x)) factor(x) else x)
adult_phylo_eco <-  adult_phylo_eco[tree.adult.eco$tip.label, ]

#make comparative object
adult.eco.comp <- comparative.data(phy=tree.adult.eco, data=adult_phylo_eco, names.col=genus_species, vcv=TRUE, na.omit=FALSE, warn.dropped=TRUE)
```

#### Anisotropy 

Here we fit a model for anisotropy ~ habitat

```{r}
#pgls for anisotropy----

#fit model
eco.aniso.pgls <- pgls(mean_anisotropy.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(eco.aniso.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(eco.aniso.pgls, which="lambda"))

#view main effects
anova(eco.aniso.pgls)

#view coefficients
summary(eco.aniso.pgls)
```

In this model the comparison state was aquatic species; to get pairwise comparisons for all groups we need to run this setting each group as the comparison state. 

Fossorial as comparison state:

```{r}
#make fossorial base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("fossorial", "ground-dwelling", "scansorial", "semi-aquatic", "subfossorial","aquatic"))

#fit model with fossorial comparison state
eco.aniso.pgls2 <- pgls(mean_anisotropy.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.aniso.pgls2)
```

Semiaquatic as comparison state:

```{r}
#make semiaquatic base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("semi-aquatic", "subfossorial","aquatic", "fossorial","ground-dwelling", "scansorial"))

#fit model with fossorial comparison state
eco.aniso.pgls2 <- pgls(mean_anisotropy.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.aniso.pgls2)
```

Subfossorial as comparison state:

```{r}
#make subfossorial base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("subfossorial","aquatic", "fossorial","ground-dwelling", "scansorial","semi-aquatic"))

#fit model with fossorial comparison state
eco.aniso.pgls2 <- pgls(mean_anisotropy.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.aniso.pgls2)
```

Ground-dwelling as comparison state:

```{r}
#make ground-dwelling base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("ground-dwelling", "scansorial","semi-aquatic","subfossorial","aquatic", "fossorial"))

#fit model with fossorial comparison state
eco.aniso.pgls2 <- pgls(mean_anisotropy.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.aniso.pgls2)
```

#### Sphericity

Here we fit a model for sphericity ~ habitat

```{r}

#pgls for sphericity----

#make aquatic base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("aquatic", "fossorial","semi-aquatic","subfossorial","ground-dwelling", "scansorial"))

#fit model
eco.sph.pgls <- pgls(mean_sphericity.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(eco.sph.pgls)
par(mfrow = c(1, 1))

#plot likelihood profile of lambda
plot(pgls.profile(eco.sph.pgls, which="lambda"))

#view main effects
anova(eco.sph.pgls)

#view coefficients
summary(eco.sph.pgls)
```

In this model the comparison state was aquatic species; to get pairwise comparisons for all groups we need to run this setting each group as the comparison state. 

Fossorial as comparison state:

```{r}

#make fossorial base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("fossorial","semi-aquatic","subfossorial","ground-dwelling", "scansorial","aquatic"))

#fit model
eco.sph.pgls <- pgls(mean_sphericity.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.sph.pgls)
```

Semiaquatic as comparison state:

```{r}

#make semiaquatic base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("semi-aquatic","subfossorial","ground-dwelling", "scansorial","aquatic", "fossorial"))

#fit model
eco.sph.pgls <- pgls(mean_sphericity.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#view coefficients
summary(eco.sph.pgls)
```

Subfossorial as comparison state:

```{r}
#make subfossorial base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("subfossorial","ground-dwelling", "scansorial", "aquatic", "fossorial","semi-aquatic"))

#fit model
eco.sph.pgls <- pgls(mean_sphericity.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(eco.sph.pgls)
par(mfrow = c(1, 1))

#plot likelihood profile of lambda
plot(pgls.profile(eco.sph.pgls, which="lambda"))

#view main effects
anova(eco.sph.pgls)

#view coefficients
summary(eco.sph.pgls)
```

Ground-dwelling as comparison state:

```{r}
#make ground-dwelling base state
adult.eco.comp$data$adult_ecology <- factor(adult.eco.comp$data$adult_ecology, levels = c("ground-dwelling", "scansorial", "aquatic", "fossorial","semi-aquatic","subfossorial"))

#fit model
eco.sph.pgls <- pgls(mean_sphericity.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(eco.sph.pgls)
par(mfrow = c(1, 1))

#plot likelihood profile of lambda
plot(pgls.profile(eco.sph.pgls, which="lambda"))

#view main effects
anova(eco.sph.pgls)

#view coefficients
summary(eco.sph.pgls)
```

#### Elongation

```{r}
#pgls for elongation----

#fit model
eco.elong.pgls <- pgls(mean_elongation.adults ~ adult_ecology,
                       data = adult.eco.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(eco.elong.pgls)
par(mfrow = c(1, 1))

#plot likelihood profile of lambda
plot(pgls.profile(eco.elong.pgls, which="lambda"))

#view main effects
anova(eco.elong.pgls)

#view coefficients
summary(eco.elong.pgls)
```

### Activity period

Here, we test for associations between activity period and lens shape among adult anurans. 

```{r}
#pgls regressions testing for effect of adult activity period on each 3D shape metric (separately)

#prep data and tree----

#subset adults with activity period data
adult_phylo_activ <- filter(adult_phylo_means, !is.na(activity_period))
table(adult_phylo_activ$activity_period)

#trim tree to adults with activity period data
rownames(adult_phylo_activ) <- adult_phylo_activ$genus_species;
name.check(tree.full, adult_phylo_activ)->overlap
#overlap
drop.tip(tree.full, overlap$tree_not_data) -> tree.adult.activ;
#plot(tree.adult.activ)

#Remove excess data from species not found in tree
adult_phylo_activ<-adult_phylo_activ[ ! adult_phylo_activ$genus_species %in% overlap$data_not_tree, ]
adult_phylo_activ[] <- lapply(adult_phylo_activ, function(x) if(is.factor(x)) factor(x) else x)
adult_phylo_activ <-  adult_phylo_activ[tree.adult.activ$tip.label, ]

#make comparative object
adult.act.comp <- comparative.data(phy=tree.adult.activ, data=adult_phylo_activ, names.col=genus_species, vcv=TRUE, na.omit=FALSE, warn.dropped=TRUE)

```

##### Anisotropy

```{r}

#pgls for anisotropy----

#make nocturnal base state
adult.act.comp$data$activity_period <- factor(adult.act.comp$data$activity_period, levels = c("nocturnal", "diurnal","both"))

#fit model
act.aniso.pgls <- pgls(mean_anisotropy.adults ~ activity_period,
                       data = adult.act.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(act.aniso.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(act.aniso.pgls, which="lambda"))

#view main effects
anova(act.aniso.pgls)

#view coefficients
summary(act.aniso.pgls)

```

#### Sphericity

```{r}

#pgls for sphericity----

#fit model
act.sph.pgls <- pgls(mean_sphericity.adults ~ activity_period,
                       data = adult.act.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(act.sph.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(act.aniso.pgls, which="lambda"))

#view main effects
anova(act.sph.pgls)

#view coefficients
summary(act.sph.pgls)
```

##### Elongation

```{r}
#pgls for elongation----

#fit model
act.elong.pgls <- pgls(mean_elongation.adults ~ activity_period,
                       data = adult.act.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(act.elong.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(act.aniso.pgls, which="lambda"))

#view main effects
anova(act.elong.pgls)

#view coefficients
summary(act.elong.pgls)
```

### Procrustes MANOVA: Adult lens shape vs. ecology

#### Adult habitat

```{r, eval=FALSE}

#procD.pgls of adult lens shape (all 3 measures) by habitat
procD<-procD.pgls(adult_phylo_eco[,c("mean_anisotropy.adults","mean_elongation.adults","mean_sphericity.adults")]~adult_ecology, tree.adult.eco, Cov = NULL, iter = 999,seed = NULL, int.first = FALSE, SS.type = NULL, effect.type = NULL, data = adult_phylo_eco, print.progress = TRUE)

summary(procD)
```

#### Activity period

```{r, eval=FALSE}
#procD.pgls of adult eye shape (all 3 measures) by activity period

procD<-procD.pgls(adult_phylo_activ[,c("mean_anisotropy.adults","mean_elongation.adults","mean_sphericity.adults")]~activity_period, tree.adult.activ, Cov = NULL, iter = 999,seed = NULL, int.first = FALSE, SS.type = NULL, effect.type = NULL, data = adult_phylo_activ, print.progress = TRUE)

summary(procD)

```

### PGLS: Adult lens size vs. eye size + ecology

Here, we test whether lens size (relative to eye size) varies across habitats or activity periods among adult anurans (n=81 species) using PGLS regressions in geomorph.  

#### Habitat

```{r}

# Prep data and tree-----

#view distribution of data
table(adult_eye$adult_ecology)

# Make tree-trimmed dataset for habitat analysis of lens size 
adult_eye_eco <- filter(adult_eye, !is.na(adult_ecology)) %>% droplevels()
adult_eye_eco[] <- lapply(adult_eye_eco, function(x) if(is.factor(x)) factor(x) else x)
adult_eye_eco<-as.data.frame(adult_eye_eco)

#trim tree to adult data
rownames(adult_eye_eco) <- adult_eye_eco$genus_species;
name.check(tree.full, adult_eye_eco) -> overlap
#overlap
drop.tip(tree.full, overlap$tree_not_data) -> tree_adult_eye_eco
#plot(tree_adult_eye_eco)

#Remove excess data from species not found in tree
adult_eye_eco<-adult_eye_eco[ ! adult_eye_eco$genus_species %in% overlap$data_not_tree, ]
adult_eye_eco[] <- lapply(adult_eye_eco, function(x) if(is.factor(x)) factor(x) else x)

#sort data to match phylo order
adult_eye_eco <-  adult_eye_eco[tree_adult_eye_eco$tip.label, ]

#make comparative object
hab.eye.comp <- comparative.data(phy = tree_adult_eye_eco, 
                                 data = adult_eye_eco, 
                                 names.col = genus_species, 
                                 vcv=TRUE, 
                                 na.omit = FALSE, 
                                 warn.dropped = TRUE)

#make aquatic comparison state
hab.eye.comp$data$adult_ecology <- factor(hab.eye.comp$data$adult_ecology, levels = c("aquatic", "fossorial","semi-aquatic", "subfossorial", "ground-dwelling", "scansorial"))


#run PGLS using caper-----

# pgls in caper of lens ~ eye for plotting residuals---


#fit pgls model for lens vs. eye size
hab.eye.pgls <- pgls(mean.lens.diameter ~ mean.eye.diameter + adult_ecology,
                       data = hab.eye.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(hab.eye.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(hab.eye.pgls, which="lambda"))

#view main effects
anova(hab.eye.pgls)

#view coefficients
summary(hab.eye.pgls)

```
To get pairwise comparisons across all habitats, we have to run with different states as comparison state. 

Fossorial as comparison state:

```{r}
#make fossorial comparison state
hab.eye.comp$data$adult_ecology <- factor(hab.eye.comp$data$adult_ecology, levels = c("fossorial","semi-aquatic", "subfossorial", "ground-dwelling", "scansorial","aquatic"))

#fit pgls model for lens vs. eye size
hab.eye.pgls2 <- pgls(mean.lens.diameter ~ mean.eye.diameter + adult_ecology,
                       data = hab.eye.comp, 
                       lambda="ML")

#view coefficients
summary(hab.eye.pgls2)
```

Semiaquatic as comparison state:

```{r}
#make semiaquatic comparison state
hab.eye.comp$data$adult_ecology <- factor(hab.eye.comp$data$adult_ecology, levels = c("semi-aquatic", "subfossorial", "ground-dwelling", "scansorial","aquatic","fossorial"))

#fit pgls model for lens vs. eye size
hab.eye.pgls2 <- pgls(mean.lens.diameter ~ mean.eye.diameter + adult_ecology,
                       data = hab.eye.comp, 
                       lambda="ML")

#view coefficients
summary(hab.eye.pgls2)
```

Subfossorial as comparison state:

```{r}
#make semiaquatic comparison state
hab.eye.comp$data$adult_ecology <- factor(hab.eye.comp$data$adult_ecology, levels = c("subfossorial", "ground-dwelling", "scansorial","aquatic","fossorial","semi-aquatic"))

#fit pgls model for lens vs. eye size
hab.eye.pgls2 <- pgls(mean.lens.diameter ~ mean.eye.diameter + adult_ecology,
                       data = hab.eye.comp, 
                       lambda="ML")

#view coefficients
summary(hab.eye.pgls2)
```

Ground-dwelling as comparison state:

```{r}
#make ground-dwelling comparison state
hab.eye.comp$data$adult_ecology <- factor(hab.eye.comp$data$adult_ecology, levels = c("ground-dwelling", "scansorial","aquatic","fossorial","semi-aquatic","subfossorial"))

#fit pgls model for lens vs. eye size
hab.eye.pgls2 <- pgls(mean.lens.diameter ~ mean.eye.diameter + adult_ecology,
                       data = hab.eye.comp, 
                       lambda="ML")

#view coefficients
summary(hab.eye.pgls2)
```

#### Activity period

```{r}
# Prep data and tree-----

#view distribution of data
table(adult_eye$activity_period)

#Make tree-trimmed dataset for activity analysis of lens size 
adult_eye_activ<-filter(adult_eye, !is.na(activity_period)) %>% droplevels()
adult_eye_activ[] <- lapply(adult_eye_activ, function(x) if(is.factor(x)) factor(x) else x)
table(adult_eye_activ$activity_period)
adult_eye_activ<-as.data.frame(adult_eye_activ)

#trim tree to adult data
rownames(adult_eye_activ) <- adult_eye_activ$genus_species;
name.check(tree.full, adult_eye_activ) -> overlap
#overlap
drop.tip(tree.full, overlap$tree_not_data)->tree.eye.activ;
#plot(tree.eye.activ)

#Remove excess data from species not found in tree
adult_eye_activ<-adult_eye_activ[ ! adult_eye_activ$genus_species %in% overlap$data_not_tree, ]
adult_eye_activ[] <- lapply(adult_eye_activ, function(x) if(is.factor(x)) factor(x) else x)
adult_eye_activ <-  adult_eye_activ[tree.eye.activ$tip.label, ]

#different factor order for contrasts
adult_eye_activ$activity_period <- factor(adult_eye_activ$activity_period, levels = c("nocturnal", "diurnal", "both"))

#make comparative object
act.eye.comp <- comparative.data(phy=tree.eye.activ, data=adult_eye_activ, names.col=genus_species, vcv=TRUE, na.omit=FALSE, warn.dropped=TRUE)

#run PGLS using caper-----

# pgls in caper of lens ~ eye 

#fit pgls model for lens vs. eye size
act.eye.pgls <- pgls(mean.lens.diameter ~ mean.eye.diameter + activity_period,
                       data = act.eye.comp, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(act.eye.pgls)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(act.eye.pgls, which="lambda"))

#view main effects
anova(act.eye.pgls)

#view coefficients
summary(act.eye.pgls)
```
Alternative activity period analyses with diurnal + neither grouped together to increase statistical power (both have small sample sizes compared to nocturnal). 

```{r}

#combine diurnal and both into one category
adult_eye_activ <- adult_eye_activ %>%
  mutate(act_bin = recode_factor(activity_period, "nocturnal" = "nocturnal", "diurnal" = "non-nocturnal", "both"="non-nocturnal"))

#view distribution of data
table(adult_eye_activ$act_bin)


#run PGLS using caper-----

# pgls in caper of lens ~ eye 

#make comparative object
act.eye.comp.bin <- comparative.data(phy=tree.eye.activ, data=adult_eye_activ, names.col=genus_species, vcv=TRUE, na.omit=FALSE, warn.dropped=TRUE)

#fit pgls model for lens vs. eye size
act.eye.pgls.bin <- pgls(mean.lens.diameter ~ mean.eye.diameter + act_bin,
                       data = act.eye.comp.bin, 
                       lambda="ML")

#check model assumptions
par(mar = c(4,4,2,2))
par(mfrow = c(2, 2))
plot(act.eye.pgls.bin)
par(mfrow = c(1, 1))

#look at phy signal estmiate
plot(pgls.profile(act.eye.pgls.bin, which="lambda"))

#view main effects
anova(act.eye.pgls.bin)

#view coefficients
summary(act.eye.pgls.bin)
```

# Figures

```{r}
#define a colorblind-friendly vector of colors for adult habitat
col_hab <- c("aquatic" = "#0072B2",
              "fossorial" = "#D55E00",
              "ground-dwelling" = "#E69F00",
              "scansorial" = "#009E73",
              "semi-aquatic" = "#56B4E9",
              "subfossorial" = "#CC79A7")

#create vector of colors for activity period
col_act <- c("both" = "deeppink3",
             "diurnal" = "darkgoldenrod1", 
             "nocturnal" = "blueviolet")

#create dataframe of artifical lens shapes
lens_sh <- specimen_means %>%
  filter(life_stage=="artificial") %>%
  add_row(specimen_id = "Perfect sphere", mean_anisotropy = 0, mean_flatness = 1, mean_sphericity = 1, mean_elongation = 1) %>%
  filter(specimen_id != "Sphere")
```


## Figure 4: Adult lens shape by ecology

Part 1: PCA centroids for adult lens shape vs ecology (habitat and activity period)

```{r Figure 4ab}

#Prep data

#pull out adult data
adults_mean <- filter(species_stage_means, life_stage == "adult") %>% droplevels()

#filter species missing habitat data
adult_phylo_eco<-filter(adults_mean, adult_ecology != "unknown")
#table(adult_phylo_eco$adult_ecology)

#add rownames 
rownames(adult_phylo_eco)<-adult_phylo_eco$genus_species;
adult_phylo_eco[] <- lapply(adult_phylo_eco, function(x) if(is.factor(x)) factor(x) else x)
rownames(adult_phylo_eco)<-adult_phylo_eco$genus_species;

#filter species missing activity period data
adult_phylo_activ<-filter(adults_mean, activity_period != "unknown")
#table(adult_phylo_activ$activity_period)

#add rownames
rownames(adult_phylo_activ)<-adult_phylo_activ$genus_species;
adult_phylo_activ[] <- lapply(adult_phylo_activ, function(x) if(is.factor(x)) factor(x) else x)
rownames(adult_phylo_activ)<-adult_phylo_activ$genus_species;

#Habitat-----

#PCA
eco.pca <- prcomp(adult_phylo_eco[,c("mean_anisotropy","mean_elongation","mean_sphericity")], center = TRUE, scale. = TRUE)

Eco <- as.factor(adult_phylo_eco$adult_ecology)
names(Eco)<-row.names(adult_phylo_eco)
EcoColr<-c("#0072B2","#D55E00","#E69F00","#009E73","#56B4E9","#CC79A7","white")
names(EcoColr) <- levels(Eco)
EcoColr<- EcoColr[match(Eco, names(EcoColr))] # col.gp must NOT be a factor

ecoPlot<-pca2d(eco.pca$x, 
               group= Eco, 
               shape = 16, 
               col =EcoColr, 
               show.centroids = TRUE, 
               radius = 1.5)

legend("bottomright", ecoPlot$groups, col=ecoPlot$colors, pch=ecoPlot$pch, box.lty=0)

# Activity Pattern ------

#PCA
activ.pca <- prcomp(adult_phylo_activ[,c("mean_anisotropy","mean_elongation","mean_sphericity")], center = TRUE, scale. = TRUE)

activ <- as.factor(adult_phylo_activ$activity_period)
names(activ)<-row.names(adult_phylo_activ)
activColr<-c("#CD1076","#FFB90F","#8A2BE2")
#activColr<-c(values=(magma(3, direction = -1)))
names(activColr) <- levels(activ)
activColr<- activColr[match(activ, names(activColr))] # col.gp must NOT be a factor

activPlot<-pca2d(activ.pca$x, 
                 group= activ, 
                 shape = 16, 
                 col =activColr, 
                 show.centroids = TRUE, 
                 radius =1.5)

legend("bottomright", activPlot$groups, col=activPlot$colors, pch=activPlot$pch, box.lty=0)

#export panels A & B-----

#print graph
pdf(file = "../Outputs/Figures/Fig4AB.pdf", width = 13,  height = 5, useDingbats=FALSE)
par(mfrow=c(1,2))

ecoPlot<-pca2d(eco.pca$x, group= Eco, shape = 16 , col =EcoColr, show.centroids = TRUE, radius =1.5)
legend("bottomright", ecoPlot$groups, col=ecoPlot$colors, pch=ecoPlot$pch, box.lty=0)

activPlot<-pca2d(activ.pca$x, group= activ, shape = 16 , col =activColr, show.centroids = TRUE, radius =1.5)
legend("bottomright", c("diurnal", "nocturnal", "neither"), col=c("#FFB90F", "#8A2BE2", "#CD1076"), pch=16, box.lty=0)

dev.off()
```

Part 2: Violin plots of adult lens shape across ecology (habitat and activity period)

```{r Figure 4cdef}

# Adult habitat ------

#anisotropy
anis_habs <- ggplot(adult_phylo_eco, aes(x=adult_ecology, y=mean_anisotropy, fill=adult_ecology)) + 
  geom_violin(trim=FALSE, alpha = 0.7)+
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  labs(title=NULL,x="Habitat", y = "Anisotropy value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic() +
  ylim(c(0, 0.6)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_anisotropy, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
 
#sphericity
sph_habs <- ggplot(adult_phylo_eco, aes(x=adult_ecology, y=mean_sphericity, fill=adult_ecology)) + 
  geom_violin(trim=FALSE, alpha = 0.7)+
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  labs(title=NULL,x="Habitat", y = "Sphericity value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic() +
  ylim(c(0.85, 1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_sphericity, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
 

#elongation
elong_habs <- ggplot(adult_phylo_eco, aes(x=adult_ecology, y=mean_elongation, fill=adult_ecology)) + 
  geom_violin(trim=FALSE, alpha = 0.7)+
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  labs(title=NULL,x="Habitat", y = "Elongation value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic()+
  ylim(c(0.8, 1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_elongation, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")

# Adult activity period ------
#anisotropy
anis_act <- ggplot(adult_phylo_activ, aes(x=activity_period, y=mean_anisotropy, fill=activity_period)) + 
  geom_violin(trim=FALSE, alpha = 0.5)+
  scale_fill_manual(values = col_act, name = "Adult activity period", guide = "none") +
  labs(title=NULL,x="Activity Period", y = "Anisotropy value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic() +
  ylim(c(0, 0.6)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_anisotropy, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
 
#sphericity
sph_act <- ggplot(adult_phylo_activ, aes(x=activity_period, y=mean_sphericity, fill=activity_period)) + 
  geom_violin(trim=FALSE, alpha = 0.5)+
  scale_fill_manual(values = col_act, name = "Adult activity period", guide = "none") +
  labs(title=NULL,x="Activity Period", y = "Sphericity value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic()+
  ylim(c(0.85, 1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_sphericity, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
  

#elongation
elong_act <- ggplot(adult_phylo_activ, aes(x=activity_period, y=mean_elongation, fill=activity_period)) + 
  geom_violin(trim=FALSE, alpha = 0.5)+
  scale_fill_manual(values = col_act, name = "Adult activity period") +
  labs(title=NULL,x="Activity Period", y = "Elongation value")+
  geom_boxplot(width=0.15, fill="white", outlier.alpha = 0)+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic() +
  ylim(c(0.8, 1)) +
  geom_hline(data = lens_sh, aes(yintercept = mean_sphericity, linetype = specimen_id)) +
  scale_linetype_manual(values = c("dashed","dotted"),
                        breaks = c("Perfect sphere", "Lens_replica"),
                        labels= c("Perfect sphere","Lens replica"),
                        name = "Reference shapes")
 

# Construct main figure -----

#make panels without lengends
fig.c <- anis_habs + 
  theme(legend.position = "none", 
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        text = element_text(size = 16)) +
  ylab("Anisotropy") +
  scale_x_discrete(breaks =c ("aquatic", "fossorial", "semi-aquatic", "subfossorial", "ground-dwelling", "scansorial"),
                   limits = c("aquatic", "fossorial", "semi-aquatic", "subfossorial", "ground-dwelling", "scansorial"),
                   labels = c("aq.", "foss.", "semiaq.", "subfoss.", "ground", "scans.")) +
  geom_text(data = data.frame("adult_ecology" = c("fossorial","aquatic","semi-aquatic","subfossorial","ground-dwelling","scansorial"),"y" = c(0.5,0.5,0.5,0.5,0.5,0.5), "group" = c("a", "a", "b", "b","b","b")),
            aes(y = y, label = group),
            vjust = -2.2)
 
 
fig.d <- anis_act + 
  theme(legend.position = "none",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        text = element_text(size = 16)) +
  scale_x_discrete(breaks = c("diurnal","nocturnal","both"),
                   limits = c("diurnal","nocturnal","both"),
                   labels = c("diurnal", "nocturnal", "other"))

fig.e <- sph_habs + 
  theme(legend.position = "none",
        text = element_text(size = 16)) +
  xlab("Habit") +
  ylab("Sphericity") +
  scale_x_discrete(breaks = c("fossorial","aquatic","semi-aquatic", "subfossorial","ground-dwelling","scansorial"),
                   limits = c("fossorial","aquatic","semi-aquatic", "subfossorial","ground-dwelling","scansorial"),
                   labels = c("foss.","aq.","semiaq.","subfoss.", "ground", "scans.") ) +
  geom_text(data = data.frame("adult_ecology" = c("fossorial","aquatic","semi-aquatic","subfossorial","ground-dwelling","scansorial"),"y" = c(1,1,1,1,1,1), "group" = c("a,b", "a,b,c", "a,b,c,d", "a,c,d","c,d","c,d")),
            aes(y = y, label = group),
            vjust = -.5)

  
fig.f <- sph_act + 
  theme(legend.position = "none",
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        text = element_text(size = 16)) +
  scale_x_discrete(breaks = c("diurnal","nocturnal","both"),
                   limits = c("diurnal","nocturnal","both"),
                   labels = c("diur.", "noct.", "neith.")) 

leg <- get_legend(sph_act + theme(legend.position="bottom"))


#make panels just with anisotropy and sphericity for habitat and activity period
plots <- plot_grid(fig.c, fig.d, fig.e, fig.f,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("C", "D", "E", "F"), #panel labels for figure
           hjust = 0, #adjustment for panel labels
           nrow = 2,
           rel_widths = c(1.8,1))

#add legend
plots2 <- plot_grid(plots, leg,
                    align = 'vh',
                    nrow = 2,
                    rel_heights = c(1, 0.1))

#print figure
plots2

#export figure
pdf(file = "../Outputs/Figures/Fig4cdef.pdf", width = 8,  height = 8)
plots2
dev.off()

# Supplemental figure -----

#make panels with elongation for habitat and activity period
plots <- plot_grid(elong_habs + theme(legend.position = "none"), 
                   elong_act + theme(legend.position = "none"),
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A", "B"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 1,
           rel_widths = c(1.8,1))

#add legend
plots2 <- plot_grid(plots, leg,
                    align = 'vh',
                    nrow = 2,
                    rel_heights = c(1, 0.1))

#print figure
plots2

#export figure
pdf(file = "../Outputs/Figures/Fig-supp-adultelong.pdf", width = 10,  height = 4)
plots2
dev.off()
```


## Figure 5: Violins of relative lens size by microhabitat (Panel A) and activity pattern (Panel B)

```{r Figure 5}

#### Habitat ####

# pgls in caper of lens ~ eye for plotting residuals---

#fit pgls model for lens vs. eye size
hab.lenseye.pgls <- pgls(mean.lens.diameter ~ mean.eye.diameter,
                       data = hab.eye.comp, 
                       lambda="ML")


#extract pgls residuals 
pglsres.habeye <- residuals(hab.lenseye.pgls) 

#name residuals
colnames(pglsres.habeye) <- "pglsres.habeye" 

#merge residuals with original data by rowname
adult_eye_eco2 <- merge(adult_eye_eco, pglsres.habeye, by = "row.names")

#plot with caper pgls residuals
gg_eco <- ggplot(adult_eye_eco2, aes(x=adult_ecology, y=pglsres.habeye, fill=adult_ecology)) + 
  geom_violin(trim=FALSE, alpha = 0.7)+
  scale_fill_manual(values = col_hab, name = "Adult habitat") +
  labs(title=NULL,x="Habitat", y = "Residual Lens Size")+
  geom_boxplot(width=0.15, fill="white")+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic()  +
  ylim(c(-1.6, 1.2))

#### Activity Pattern ####

# pgls in caper of lens ~ eye for plotting residuals---

#fit pgls model for lens vs. eye size
act.lenseye.pgls <- pgls(mean.lens.diameter ~ mean.eye.diameter,
                       data = act.eye.comp, 
                       lambda="ML")

#extract pgls residuals 
pglsres.acteye <- residuals(act.lenseye.pgls) 

#name residuals
colnames(pglsres.acteye) <- "pglsres.acteye" 

#merge residuals with original data by rowname
adult_eye_activ2 <- merge(adult_eye_activ, pglsres.acteye, by = "row.names")

#make plot of caper pgls resudals across activity periods
gg_activ <- ggplot(adult_eye_activ2, aes(x=activity_period, y=pglsres.acteye, fill=activity_period)) + 
  geom_violin(trim=FALSE, alpha = 0.5)+
  scale_fill_manual(values = col_act, name = "Adult activity period") +
  labs(title=NULL,x="Activity Period", y = "Residual Lens Size")+
  geom_boxplot(width=0.15, fill="white")+
  geom_jitter(shape = 19, size = 1, alpha = 0.7, position = position_jitter(0.1)) + 
  theme_classic() +
  ylim(c(-1.6, 1.2))

gg_activ 

#Construct figure----

fig.a <- gg_eco + 
  theme(legend.position = "none", 
        text = element_text(size = 16)) +
  scale_x_discrete(breaks = c("aquatic","fossorial","semi-aquatic", "subfossorial","ground-dwelling","scansorial"),
                   limits = c("aquatic","fossorial","semi-aquatic", "subfossorial","ground-dwelling","scansorial"),
                   labels = c("aq.","foss.","semiaq.","subfoss.", "ground", "scans.") ) +
  xlab("Habit")
  
  
fig.b <- gg_activ + 
  theme(legend.position = "none", 
        text = element_text(size = 16)) +
  scale_x_discrete(breaks = c("diurnal","nocturnal","both"),
                 limits = c("diurnal","nocturnal","both"),
                 labels = c("diur.", "noct.", "neith."))

plots <- plot_grid(fig.a, fig.b,
           align = 'vh', #align horizontally and vertically
           axis = 'lb',
           labels = c("A", "B"), #panel labels for figure
           hjust = -.1, #adjustment for panel labels
           nrow = 1,
           rel_widths = c(1.8,1))

plots

#print graph
pdf(file = "Figures/Fig5.pdf", width = 8,  height = 4)
plots
dev.off()

```


